Characterizing the response of subarctic European vegetation to key environmental variables for land surface modeling of ozone pollution and risk assessment

### 2021-10-08 VegOzone.tex
Restructure conclusions

To summarize, we have studied the effect of bespoke parameterizations of common vegetation types (coniferous and deciduous trees, perennial grassland) to a subarctic climate on $\mathrm{POD_y}$. The comparison between meteorological conditions in 2018 and 2019 and their divergence from climatology allowed us to assess the influence of key environmental variables such as temperature, PPFD, and precipitation on vegetation susceptibility to \chem{O_3} damage in light of future changes as may occur under climate change. We found that conditions for ozone formation were more favorable in the 2018 growing season than in 2019, with 2018 being significantly warmer and less cloudy in spring and early summer. Accordingly, peak ozone concentrations occurred more frequently and at higher levels in 2018. This was particularly a result of the extended heatwave in spring and early summer and associated, extensive wildfires in Central Sweden. However, ozone concentrations during the growing season were not significantly enhanced and \chem{POD_1} values were very similar for both years.

Based on a MODIS photosynthesis product \citep{MODIS_PSN}, we estimated the growing season for coniferous trees in 2018 to be at least one month shorter than in 2019. The determined start of growing season in 2018 ($\unit{doy}~122$) and 2019 ($\unit{doy}~106$) lies within the range given by observations in Fennoscandia \citep{TB:Kolari2007,IVL:Karlsson2018}. We deduced the start of the growing season for deciduous trees based on gridded observational temperature and for perennial grassland on snow depth data. For perennial grassland, we assumed a one month latency period after snow melt which is supported by observational data from Finland \citep{FCR:Korhonen2018}. With respect to ongoing climate change, a clear positive trend emerged in length ($5.2\,\unit{d\,decade^{-1}}$) of the growing season which is almost equally distributed between earlier start ($2.9\,\unit{days\,decade^{-1}}$) and later end ($2.3\,\unit{d\,decade^{-1}}$) (Appendix Fig.~\ref{fig:greening_season_change_Svanvik}).

This suggests that the overall length of the growing season is more crucial in risk assessment when using flux-based metrics than episodes of peak concentrations as these peak \chem{[O_3]} often coincide with environmental conditions that limit ozone uptake in the $\mathrm{DO_3SE}$ model. Visible damage caused by peak ozone uptake and a dependence of ozone sensibility on phenology are not accounted for in $\mathrm{POD_y}$. But open-top chamber (OTC) experiments performed in northern Finland where Scots pine and downy birch seedlings were exposed to elevated ozone concentrations showed a reduction in biomass and reproduction with visible damage explicitly attributed to peak \chem{O_3} concentrations and fast phenological development at high growth rate \citep{Amb:Manninen2009}. In particular, forbs and perennial grassland have been found to be more susceptible to ozone-induced damage in the reproductive state \citep{EP:Bassin2004}. That might explain why we observed ozone damage to a larger extent in 2018 than in 2019. An associated damage function to ozone depending on leaf age as proposed by \citet{AE:Musselman2006} or phenological state might improve the predictions.

At the same time, intra-species variability even at the same location is typically non-negligible \citep{EP:Bassin2004,Amb:Girgzdiene2009} but not accounted for in the models. This may also explain some contradictory results regarding ozone sensitivity of natural vegetation in northern Europe. \citet{FS:Subramanian2014} reported a higher, modeled reduction of net primary production in conifers ($4.3-15.5\,\unit{\%}$) than birch ($1.4-4.3\,\unit{\%}$) under elevated ozone in Sweden, while \citet{Amb:Girgzdiene2009} observed more visible damage on deciduous trees than on Norway spruce in Lithuania. But visible damage and biomass reduction may not occur at the same rate and hence may not be interchangeable. In our study, we do not explicitly account for this natural distribution of individual plant traits, but our bespoke response functions reflect the diversity within a given species. In particular, we assumed two tolerance levels to cold temperatures leading to a higher efficiency in photosynthesis at cool temperatures than for central European species. To this end, we computed a PDF from observed temperature at the NIBIO Environment Centre Svanhovd (May--September, 1992--2000). We adjusted the parameters in the temperature response function of stomatal conductance to increase the enclosed area with the PDF. Assuming a proportionality between $g_\mathrm{sto}$ and $A_\mathrm{n}$, this is equivalent to optimizing carbon uptake. Similarly, we deduced a PPFD PDF and varied the light response function to test a higher or lower extent of stomatal opening at low light intensities. These parameterizations are hypothetical and need verification by experimental data. We found that soil water potential under 2018/19 meteorological conditions was negligible when considering the bespoke growing season. We have not assessed the effect of the plants' response to VPD in detail, but our climatological data indicated that conditions at Svanhovd are shifting towards a more VPD limited regime. Our key findings are: (1) Our bespoke parameterizations for subarctic species are better suited to describe vegetation adapted to subarctic conditions for instance allowing for stomata opening during a larger part of the growing season, (2) the parameterizations defined in the mapping manual for European regional risk assessment appear to not adequately capture the physiological responses of subarctic vegetation that are important determinants of \chem{[O_3]} susceptibility and are likely to underestimate ozone-damage risk, and (3) climatic conditions that are promoting a longer growing season are likely to increase the ozone-damage risk in flux-based metrics.

In this regard, vegetation might become more exposed to higher ozone concentrations in early phenological stages due to an overlap with ozone spring peak \citep{EP:Karlsson2007}. However, the decline of this ozone spring peak is partly caused by uptake of vegetation. It, therefore, remains unclear whether an earlier start of the growing season will increase the exposure of vegetation to ozone or lead to a compensation due to an earlier decline of the peak ozone in the future. \citet{ESPR:Hayes2021} pointed out that the highest susceptibility of vegetation to future ozone variability is likely to occur in summer when vegetation is most productive. The actual resilience of northern Fennoscandia plant communities to climate change and its predictions strongly depend on both the range of existing acclimations as well as the actual type.

Following the methods described in \citet{ICP:MappingManual2017}, we estimated biomass reductions due to ozone uptake for both 2018 and 2019 and found substantial reductions for deciduous trees ($15.5-17.4\,\unit{\%}$) and perennial grassland (above ground, $12.4-13.7\,\unit{\%}$) in both years. These reductions are higher than for the default parameterizations, $11.2\,\unit{\%}$ (deciduous) and $9.6\,\unit{\%}$ (perennial grassland). For coniferous trees, the estimated reduction in biomass differs insignificantly between our bespoke parameterizations ($2.5-3\,\unit{\%}$) and the default ($2.31\,\unit{\%}$). This indicates a higher modeled robustness to ozone damage in coniferous trees. All these results have to be treated with care because the biomass reduction functions were established based on species in less extreme climates. In particular, CLs for deciduous and coniferous trees were breach in all years and fora ll parameterizations. For perennial grassland, the CL was not breached if a bespoke GS was assumed, but the mapping manual defined CL of $10\,\unit{\%}$ loss in biomass might not be transferable to subarctic regions with considerable constrains on productivity leading to more severe economical consequences.

>>>>

To summarize, we have developed bespoke parameterizations of common vegetation types (coniferous and deciduous trees, perennial grassland) to a subarctic climate and studied their effect on $\mathrm{POD_y}$. The comparison between meteorological conditions in 2018 and 2019 and their divergence from climatology allowed us to assess the influence of key environmental variables such as temperature, PPFD, and precipitation on vegetation susceptibility to \chem{O_3} damage in light of future changes as may occur under climate change. We found that conditions for ozone formation were more favorable in the 2018 growing season than in 2019, with 2018 being significantly warmer and less cloudy in spring and early summer. Accordingly, peak ozone concentrations occurred more frequently and at higher levels in 2018. This was particularly a result of the extended heatwave in spring and early summer and associated, extensive wildfires in Central Sweden. However, ozone concentrations during the growing season were not significantly enhanced and \chem{POD_1} values were very similar for both years.
This suggests that the overall length of the growing season is more crucial in risk assessment when using flux-based metrics than episodes of peak concentrations as these peak \chem{[O_3]} often coincide with environmental conditions that limit ozone uptake in the $\mathrm{DO_3SE}$ model.

Based on a MODIS photosynthesis product \citep{MODIS_PSN}, we estimated the growing season for coniferous trees in 2018 to be at least one month shorter than in 2019. The determined start of growing season in 2018 ($\unit{doy}~122$) and 2019 ($\unit{doy}~106$) lies within the range given by observations in Fennoscandia \citep{TB:Kolari2007,IVL:Karlsson2018}. We deduced the start of the growing season for deciduous trees based on gridded observational temperature and for perennial grassland on snow depth data. For perennial grassland, we assumed a one month latency period after snow melt which is supported by observational data from Rovaniemi (Finland) \citep{FCR:Korhonen2018}. With respect to ongoing climate change, a clear positive trend emerged in length ($5.2\,\unit{d\,decade^{-1}}$) of the growing season which is almost equally distributed between earlier start ($2.9\,\unit{days\,decade^{-1}}$) and later end ($2.3\,\unit{d\,decade^{-1}}$) (Appendix Fig.~\ref{fig:greening_season_change_Svanvik}).

 The $\mathrm{DO_3SE}$ modeling results are in contrast to our observation of more visible damage in 2018 compared to 2019. However, visible damage caused by peak ozone uptake and a dependence of ozone sensibility on phenology are not accounted for in $\mathrm{POD_y}$. In this regard, open-top chamber (OTC) experiments performed in northern Finland where Scots pine and downy birch seedlings were exposed to elevated ozone concentrations attributed a reduction in biomass and reproduction with visible damage explicitly to peak \chem{O_3} concentrations and fast phenological development at high growth rate \citep{Amb:Manninen2009}. In particular, forbs and perennial grassland have been found to be more susceptible to ozone-induced damage in the reproductive state \citep{EP:Bassin2004}. 
That might explain why we observed ozone damage to a larger extent in 2018 than in 2019. An associated damage function to ozone depending on leaf age as proposed by \citet{AE:Musselman2006} or phenological state might improve the predictions.

At the same time, intra-species variability even at the same location is typically non-negligible \citep{EP:Bassin2004,Amb:Girgzdiene2009} but not accounted for in the models. This may also explain some contradictory results regarding ozone sensitivity of natural vegetation in northern Europe. \citet{FS:Subramanian2014} reported a higher, modeled reduction of net primary production in conifers ($4.3-15.5\,\unit{\%}$) than birch ($1.4-4.3\,\unit{\%}$) under elevated ozone in Sweden, while \citet{Amb:Girgzdiene2009} observed more visible damage on deciduous trees than on Norway spruce in Lithuania. But visible damage and biomass reduction may not occur at the same rate and hence may not be interchangeable.

In our study, we do not explicitly account for this natural distribution of individual plant traits, but our bespoke response functions reflect the diversity within a given species. In particular, we assumed two tolerance levels to cold temperatures leading to a higher efficiency in photosynthesis at cool temperatures than for central European species. To this end, we computed a PDF from observed temperature at the NIBIO Environment Centre Svanhovd (May--September, 1992--2000). We adjusted the parameters in the temperature response function of stomatal conductance to increase the enclosed area with the PDF. Assuming a proportionality between $g_\mathrm{sto}$ and $A_\mathrm{n}$, this is equivalent to optimizing carbon uptake. Similarly, we deduced a PPFD PDF and varied the light response function to test a higher or lower extent of stomatal opening at low light intensities. Remark that these parameterizations are hypothetical and need verification by experimental data. We found that soil water potential under 2018/19 meteorological conditions was negligible when considering the bespoke growing season. We have not assessed the effect of the plants' response to VPD in detail, but our climatological data indicated that conditions at Svanhovd are shifting towards a more VPD limited regime.

Our key findings are: 
\begin{enumerate}
\item Our bespoke parameterizations for subarctic species are better suited to describe vegetation adapted to subarctic conditions for instance allowing for stomata opening during a larger part of the growing season, 
\item the parameterizations defined in the mapping manual for European regional risk assessment appear to not adequately capture the physiological responses of subarctic vegetation that are important determinants of \chem{[O_3]} susceptibility and are likely to underestimate ozone-damage risk, and 
\item climatic conditions that are promoting a longer growing season are likely to increase the ozone-damage risk in flux-based metrics.
\end{enumerate}

In this regard, vegetation might become more exposed to higher ozone concentrations in early phenological stages due to an overlap with ozone spring peak \citep{EP:Karlsson2007}. However, the decline of this ozone spring peak is partly caused by uptake of vegetation. It, therefore, remains unclear whether an earlier start of the growing season will increase the exposure of vegetation to ozone or lead to a compensation due to an earlier decline of the peak ozone in the future. \citet{ESPR:Hayes2021} pointed out that the highest susceptibility of vegetation to future ozone variability is likely to occur in summer when vegetation is most productive. The actual resilience of northern Fennoscandia plant communities to climate change and its predictions strongly depend on both the range of existing acclimations as well as the actual type acclimation.

Following the methods described in \citet{ICP:MappingManual2017}, we estimated biomass reductions due to ozone uptake for both 2018 and 2019 and found substantial reductions for deciduous trees ($15.5-17.4\,\unit{\%}$) and perennial grassland (above ground, $12.4-13.7\,\unit{\%}$) in both years. These reductions are higher than for the default parameterizations, $11.2\,\unit{\%}$ (deciduous) and $9.6\,\unit{\%}$ (perennial grassland). For coniferous trees, the estimated reduction in biomass differs insignificantly between our bespoke parameterizations ($2.5-3\,\unit{\%}$) and the default ($2.31\,\unit{\%}$). This indicates a higher modeled robustness to ozone damage in coniferous trees. All these results have to be treated with care because the biomass reduction functions were established based on species in less extreme climates. In particular, CLs for deciduous and coniferous trees were breached in all years and forall parameterizations. For perennial grassland, the CL was not breached if a bespoke GS was assumed, but the mapping manual defined CL of $10\,\unit{\%}$ loss in biomass might not be transferable to subarctic regions with considerable constrains on productivity leading to more severe economical consequences.
